---
title: "Untitled"
author: "Mphaphuli M.P"
date: "2023-11-13"
output:
  word_document: default
  html_document: default
---

# Loading libraries
```{r}
library(geoR)
library(gstat)
library(fields)
```

# Observation Point + Observation region
```{r}
points(sic.100, borders = sic.borders, col = "darkblue")
points(sic.367, borders = sic.borders, col = "lightblue", add = TRUE)
legend("topright", legend = c("sic.100", "sic.367"), col = c("darkblue", "lightblue"), pch = 1)
```
```{r}
plot.geodata(sic.100,bor=sic.borders)
```

```{r}
vario.b<- variog(sic.100,option =c ("bin", "cloud", "smooth"),
                 bin.cloud=TRUE)
vario.c <- variog(sic.100, op="cloud")
bplot.xy(vario.c$u,vario.c$v, breaks=vario.b$u,col="grey80",
         lwd=2,cex=0.1,outline=FALSE)
```
# Experimental variogram
```{r}
vario.ex<- variog(sic.100, bin.cloud=TRUE)
plot(vario.ex)
```

# Directional variograms
```{r}
vario4<-variog4(sic.100)
plot(vario4,same=FALSE)
plot(variog4(sic.100, trend = "2nd", max.dist = 309.75), omnidirectional = T)
```

# Three examples of experimental variogram adjustment
```{r}
vario.ex<- variog(sic.100,option="bin")
vario.sphe<-(variofit(vario.ex,cov.model= "spher",
                      ini.cov.pars=c(15000,200)))
par(mfrow=c(2,2), mar=c(3,3,1,1), mgp =c (2,1,0))
plot(vario.ex,main="Spherical")
lines.variomodel(cov.model="sphe",cov.pars=c(15000,100),
                 nug=0,max.dist=350)
plot(vario.ex,main="Exponential")
lines.variomodel(cov.model="exp",cov.pars=c(15000,100),
                 nug=0,max.dist=350)
plot(vario.ex,main="Exponential with nugget effect")
lines.variomodel(cov.model="exp",cov.pars=c(10000,100),
                 nug=5000,max.dist=350)
plot(vario.ex,main="Exponential with nugget effect")
lines.variomodel(cov.model="matern",cov.pars=c(10000,100),
                 nug=0,max.dist=350,kappa=0.5)
```

# Spherical variogram for raw data
```{r}
vario.ex<- variog(sic.100, bin.cloud=TRUE)
plot(vario.ex,main="")
lines.variomodel(cov.model="spher",cov.pars=c(15000,50),
                 nug=0,max.dist=300)
# Extract parameters
nugget <- vario.ex$nugget
sill <- vario.ex$sill
range_param <- vario.ex$range

# Print the results
cat("Nugget:", nugget, "/n")
cat("Sill:", sill, "/n")
cat("Range:", range_param, "/n")
```

# Kriging estimates and variance for raw data
```{r}
pred.grid <- expand.grid(seq(0,350, l=51),seq (0,220, l=51))
rgb.palette <- colorRampPalette(c("darkblue", "blue", "purple", "yellow"),space = "rgb")
kc<- krige.conv(sic.100, loc = pred.grid,
                krige=krige.control(cov.model="spherical",cov.pars=c(15000,50)))
image(kc, loc = pred.grid,col =rgb.palette(20) ,xlab="Coord X",
      ylab="Coord Y",borders=sic.borders,main="Estimation")
legend_colors <- rgb.palette(20)
legend_breaks <- seq(min(kc$krige.var), max(kc$predict), length.out = length(legend_colors) + 1)
image.plot(x = NULL, y = NULL, z = matrix(legend_breaks, nrow = 1), col = rgb.palette(20),
           axes = FALSE, legend.only = TRUE, horizontal = FALSE, legend.width = 1, legend.shrink = 0.8,
           legend.mar = 5)
krige.var <- kc$krige.var
image(kc, krige.var,loc = pred.grid,col=rgb.palette(20),
      xlab="Coord X",ylab="Coord Y",borders=sic.borders,
      main="Kriging variance")
legend_colors <- rgb.palette(20)
legend_breaks <- seq(min(kc$krige.var), max(kc$predict), length.out = length(legend_colors) + 1)
image.plot(x = NULL, y = NULL, z = matrix(legend_breaks, nrow = 1), col = rgb.palette(20),
           axes = FALSE, legend.only = TRUE, horizontal = FALSE, legend.width = 1, legend.shrink = 0.8,
           legend.mar = 5)
```

# Estimated and observed values
```{r}
kc1<- krige.conv(sic.100, loc = sic.100$coords,
                 krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
kc2<- krige.conv(sic.100, loc = sic.367$coords,
                 krige=krige.control(cov.model="spherical",cov.pars=c(16000,47)))
plot(sic.100$data,kc1$predict,xlab="Observed",ylab="Estimated",
     main="Control sample")
abline(a=0,b=1,col="red")
plot(sic.367$data,kc2$predict,xlab="Observed",ylab="Estimated",
     main="Control")
abline(a=0,b=1,col="red")
```

# Transformation
```{r}
plot.geodata(sic.100,bor=sic.borders,lambda = 0.5)
```

# Experimental variogram and theoretical variogram after transformation
```{r}
vario.ext <- variog(sic.100, option= "bin",lambda = 0.5)
plot(vario.ext)
lines.variomodel(cov.m = "mat",cov.p = c(105,36), nug = 6.9,
                 max.dist = 300, kappa = 1, lty = 1)
vario.ext <- variog(sic.100, option= "bin",lambda = 0.5)
plot(vario.ext)
lines.variomodel(cov.m = "spherical",cov.p = c(105,36), nug = 6.9,
                 max.dist = 300, kappa = 1, lty = 1)
vario.ext <- variog(sic.100, option= "bin",lambda = 0.5)
plot(vario.ext)
lines.variomodel(cov.m = "gaussian",cov.p = c(105,36), nug = 6.9,
                 max.dist = 300, kappa = 1, lty = 1)
summary(vario.ext)

# Extract parameters
nugget <- vario.ext$nugget
sill <- vario.ext$sill
range_param <- vario.ext$range

# Print the results
cat("Nugget:", nugget, "/n")
cat("Sill:", sill, "/n")
cat("Range:", range_param, "/n")
```

# Kriging estimates and variance for transformed data
```{r}
pred.grid1 <- expand.grid(seq(0,350, l = 51), seq(0,220, l = 51))
rgb.palette <- colorRampPalette(c("darkblue", "blue", "purple", "yellow"), space = "rgb")
kct <- krige.conv(sic.100, loc = pred.grid1, krige = krige.control(cov.model = "matern", cov.pars = c(105, 36), kappa = 1, nugget = 6.9, lambda = 0.5))
image(kct, loc = pred.grid, col = rgb.palette(20),
      xlab = "Coord X", ylab = "Coord Y", borders = sic.borders,
      main = "Estimation")
legend_colors <- rgb.palette(20)
legend_breaks <- seq(min(kct$krige.var), max(kct$predict), length.out = length(legend_colors) + 1)
image.plot(x = NULL, y = NULL, z = matrix(legend_breaks, nrow = 1), col = rgb.palette(20),
           axes = FALSE, legend.only = TRUE, horizontal = FALSE, legend.width = 1, legend.shrink = 0.8,
           legend.mar = 5)
krige.var1 <- kct$krige.var
image(kct, krige.var1,loc = pred.grid1, col = rgb.palette(20),xlab = "Coord X", ylab = "Coord Y", borders = sic.borders,
      main = "Kriging Variance")
legend_colors <- rgb.palette(20)
legend_breaks <- seq(min(kct$krige.var), max(kct$predict), length.out = length(legend_colors) + 1)
image.plot(x = NULL, y = NULL, z = matrix(legend_breaks, nrow = 1), col = rgb.palette(20),
           axes = FALSE, legend.only = TRUE, horizontal = FALSE, legend.width = 1, legend.shrink = 0.8,
           legend.mar = 5)
```

# Estimated and observed values for transformed data
```{r}
kct1<- krige.conv(sic.100, loc = sic.100$coords,
                  krige=krige.control(cov.model="spherical",cov.pars=c(16000,47),
                                      kappa=1,nugget=6.9,lambda=0.5))
kct2<- krige.conv(sic.100, loc = sic.367$coords,
                  krige=krige.control(cov.model="spherical",cov.pars=c(16000,47),
                                      kappa=1,nugget=6.9,lambda=0.5))
plot(sic.100$data,kct1$predict,xlab="Observed",ylab="Estimated",
     main="Sample")
abline(a=0,b=1,col="red")
plot(sic.367$data,kct2$predict,xlab="Observed",ylab="Estimated",
     main="Control sample")
abline(a=0,b=1,col="red")
```


#
```{r}
```

# Cross Validation
```{r}
# Raw data
res1 <- sic.100$data - kc1$predict
res2 <- sic.367$data - kc2$predict


## Calculate RMSE
RMSE1 <- sqrt(mean(res1^2, na.rm =TRUE))
RMSE2 <-sqrt(mean(res2^2, na.rm =TRUE))

print(paste("RMSE for Control", RMSE1))
print(paste("RMSE for Control Sample", RMSE2))

## Calculate MAE
mae <- mean(abs(res2))

print(paste("MAE for Control Sample", mae))

## Calculate R-squared
mean_observed <- mean(sic.100$data)
ss_total <- sum((sic.100$data - mean_observed)^2)
ss_residual <- sum((res1)^2)
r_squared <- 1 - (ss_residual / ss_total)

print(paste("R-squared:", r_squared))

## Calculate R-squared
mean_observed1 <- mean(sic.367$data)
ss_total1 <- sum((sic.367$data - mean_observed1)^2)
ss_residual1 <- sum((res2)^2)
r_squared1 <- 1 - (ss_residual1 / ss_total1)

print(paste("R-squared:", r_squared1))

# Transformed
rest1 <- sic.100$data - kct1$predict
rest2 <- sic.367$data- kct2$predict

rmse_control <- sqrt(mean((sic.100$data - kct1$predict)^2))

cat("RMSE for Control:", rmse_control, "/n")

rmse_controlSample <- sqrt(mean((sic.367$data - kct2$predict)^2))

cat("RMSE for Control Sample:", rmse_controlSample, "/n")
 
## Calculate RMSE
RMSE_1 <- sqrt(mean(rest1^2, na.rm =TRUE))
RMSE_2 <- sqrt(mean(rest2^2, na.rm =TRUE))

print(paste("RMSE for Control", RMSE_1))
print(paste("RMSE for Control Sample", RMSE_2))

## Calculate MAE
mae_2 <- mean(abs(rest2))

print(paste("MAE for Control Sample", mae_2))

## Calculate R-squared
mean_observed_t1 <- mean(sic.100$data)
ss_total_t1 <- sum((sic.100$data - mean_observed_t1)^2)
ss_residual_t1 <- sum((rest1)^2)
r_squared_t1 <- 1 - (ss_residual_t1 / ss_total_t1)

print(paste("R-squared:", r_squared_t1))

## Calculate R-squared
mean_observed_t2 <- mean(sic.100$data)
ss_total_t2 <- sum((sic.100$data - mean_observed_t2)^2)
ss_residual_t2 <- sum((rest2)^2)
r_squared_t2 <- 1 - (ss_residual_t2 / ss_total_t2)

print(paste("R-squared:", r_squared_t2))
```
# 
```{r}
# Extract observed values from 'sic.100'
observed_values <- sic.100$data

# Predict using the kriging model 'kct1'
predicted_values <- predict(kct1, newdata = sic.100)

# Specify lambda for Box-Cox transformation (replace with your actual lambda)
lambda <- 0.5

# Apply Box-Cox transformation to the observed values
transformed_observed <- ifelse(lambda == 0, log(observed_values), (observed_values^lambda - 1) / lambda)

# Apply Box-Cox transformation to the predicted values
transformed_predicted <- ifelse(lambda == 0, log(kct1), (kct1^lambda - 1) / lambda)

# Calculate residuals for the transformed values
residuals_transformed <- transformed_observed - kct1

```



```{r}
shapiro.test(sic.100$data) # Raw data
```
